\section{Introduction}
\label{sec:introduction}

Generative modeling of discrete structured objects, such as graphs, 
is a central challenge across many scientific and industrial applications, 
including drug discovery and materials design \citep{jin2018junction,stokes2020deep,xie2022crystal}.  While domain-driven simulations explicitly encode physical knowledge \citep{noe2020machine}, 
they quickly become intractable due to the combinatorial complexity of the underlying data distribution \citep{polishchuk2013estimation}, 
highlighting the need for generative models that efficiently capture implicit, domain-specific patterns.

Current state-of-the-art models for unconditional graph generation predominantly rely on discrete diffusion~\citep{vignac2022digress} or continuous-time Markov chains~\citep{campbell2024generative,qinmadeira2024defog}, translating samples from simple source distributions into complex graph structures. Although effective in generating realistic unconditional samples, these methods struggle when enforcing user-specified properties or constraints, as the guidance typically occurs off-manifold, where property regressors and conditioning functions are inherently less reliable.

\Glspl{ebm}~\citep{hinton2002training, lecun2006tutorial} offer a complementary perspective that naturally addresses these limitations. \Glspl{ebm} explicitly represent the probability structure of the data distribution through a scalar energy function, enabling direct incorporation of constraints, priors, or property-based objectives at inference time without retraining~\citep{du2019implicit, du2025learning}. For graph generation, \glspl{ebm} offer particular advantages, as graph-structured constraints and domain-specific priors can be explicitly and elegantly integrated into the energy function. Historically, however, \glspl{ebm} have struggled with sample quality, primarily due to poor sampling efficiency in discrete, combinatorial domains, as exemplified by existing methods such as GraphEBM~\citep{liu2021graphebm}. Consequently, discrete \glspl{ebm} have predominantly served as scoring models for out-of-distribution detection rather than competitive generative samplers~\citep{wu2023energy,fuchsgruber2024energy}.




Energy Matching~\citep{balcerak2025energy} bridges these perspectives by building on the \gls{jko} formulation~\citep{jordan1998variational} and its first-order optimality perspective~\citep{terpin2024learning,lanzetti2024variational,lanzetti2025firstorder}, explicitly coupling transport toward the data distribution with refinement through an explicitly time-indexed sequence of sampling steps. In these continuous formulations, dynamics typically follow \glspl{ode} or \glspl{sde}, guided by a learned scalar potential that defines both the transport direction and refinement by capturing a Boltzmann-like probability structure around the data. This regularizes energies off-support while retaining the compositional advantages of \glspl{ebm}. While effective in continuous settings, explicit time-indexed dynamics are not naturally suited to discrete graph domains, where generation proceeds via local graph edits rather than continuous trajectories. Moreover, Energy Matching's objective directly relies on \gls{ot} distances, making the formulation computationally idealized for large combinatorial graph spaces.

For discrete graph generation, this motivates a formulation that removes explicit time indexing altogether, operating purely at the level of the potential energy landscape defined by the learned scalar potential. Transitions between greedy transport-like moves and local refinement would emerge implicitly from energy-dependent acceptance criteria rather than predefined schedules. Additionally, exact \gls{ot} distances can be replaced by approximate discrepancy measures induced by stochastic interpolants~\citep{albergoBuildingNormalizingFlows2023} based on minibatch couplings~\citep{albergo2024stochastic}, providing tractable guidance toward data-supported regions. Such a formulation avoids explicitly solving computationally prohibitive \gls{ot} problems, making it more suitable for discrete graph spaces.
\input{figures/overview_simple_figure}

\begin{mdframed}[hidealllines=true,backgroundcolor=BurntOrange!8]
\paragraph{Contributions.}
We introduce \glsadd{gem}\emph{\gls{gem}}\footnote{Code will be released upon acceptance.}, a state‑of‑the‑art discrete energy‑based generative model for graphs whose generation quality matches or exceeds leading discrete diffusion approaches. By explicitly defining a Gibbs distribution over graphs, \gls{gem} captures the relative likelihood structure of the graph space, naturally supports inference‑time compositional constraints, and enables computation of graph geodesics as minimum‑energy edit paths aligned with the learned data geometry implicitly encoded in the potential.

To achieve this, \gls{gem} employs a permutation‑invariant, time‑independent potential trained under a relaxed Wasserstein discrepancy geometry. Unlike discrete diffusion or continuous energy matching methods, \gls{gem}'s sampling dynamics are free of externally imposed indexing; transitions between transport and local refinement emerge naturally and solely from the learned energy landscape. In practice, sampling combines initial greedy transport‑like moves from noise toward high‑probability regions with local graph edits informed by the learned data geometry and guided by \gls{mh} acceptance.
\end{mdframed}
