\section{Preliminaries}

%Not sure if we need this
% \paragraph{Energy-based models and sampling challenges.}
% An \gls{ebm} assigns an energy $E_\theta(x)$, inducing $\pi_\theta(x)\propto \exp(-E_\theta(x))$.
% Maximum-likelihood training uses contrastive objectives with negatives from \gls{mcmc}; the energy is
% poorly posed outside the support of the target distribution and chains can mix slowly. The energy view
% is compositional in the sense that independent factors in the posterior correspond to additive energy terms:
% $p(x \mid y)\propto \underbrace{p(y \mid x)}_{\text{conditional likelihood}}\underbrace{p(x)}_{\text{prior}}$,
% where $y$ is an observation/measurement and $x$ is a sample. The conditional term is often provided by
% a discriminator/regressor; a common Gaussian data-fidelity choice is
% $p(y\mid x)\propto \exp(-\lVert y - A(x)\rVert^2/\zeta^2)$, where $A(x)$ is a forward/measurement model
% and $\zeta$ is a scalar controlling adherence to the condition. At inference, posterior sampling is
% obtained by adding the data-fidelity term (and any other constraint energies) to $E_\theta$.

% \paragraph{Optimal transport costs, stochastic interpolants, and kinetic action.}
% Given a cost $c:\mathcal{X}\times\mathcal{X}\to[0,\infty)$ and distributions $\rho_0,\rho_1$, the
% Kantorovich transport cost is
% \begin{equation}
% W_c(\rho_0,\rho_1)
% \;\coloneqq\;
% \inf_{\gamma\in\Gamma(\rho_0,\rho_1)} \mathbb{E}_\gamma[c(X_0,X_1)],
% \end{equation}
% with $\Gamma(\rho_0,\rho_1)$ the set of couplings. For $\mathcal{X}=\mathbb{R}^d$ and
% $c(x_0,x_1)=\|x_0-x_1\|^2$, $W_c=W_2^2$. Exact OT couplings are typically expensive to compute, but for
% any coupling the induced kinetic action can be evaluated as a discrepancy that upper-bounds the true
% transport cost and provides a usable geometry. Fixing a coupling $\gamma$ with $(X_0,X_1)\sim\gamma$,
% define the interpolant $I_t=(1-t)X_0+tX_1$ and velocity field
% $b_t^\gamma(x)\coloneqq\mathbb{E}[X_1-X_0\mid I_t=x]$, which induce the kinetic action
% \begin{equation}\label{eq:kinetic-action}
% \mathcal{A}(\gamma)
% \;\coloneqq\;
% \int_0^1 \int \|b_t^\gamma(x)\|^2\,\rho_t^\gamma(\mathrm{d}x)\,\mathrm{d}t.
% \end{equation}
% For quadratic cost,
% \begin{equation}\label{eq:kinetic-action-bound}
% W_2^2(\rho_0,\rho_1)\le\mathcal{A}(\gamma)\le\mathbb{E}_\gamma[\|X_1-X_0\|^2],
% \end{equation}
% with equality for optimal deterministic (Monge) couplings. A tractable coupling
% (e.g.\ minibatch matching) is therefore used, and $\mathcal{A}(\gamma)$ is treated as a restricted
% geometry that upper-bounds the true Wasserstein distance.



\paragraph{Energy matching and a learned potential.}
Energy matching learns a parameterized ($\theta$) scalar \emph{potential} ($V_\theta$) for a data space ($\mathcal{X}$), i.e., $V_\theta:\mathcal{X}\to\mathbb{R}$. The scalar potential$V_\theta$ plays two roles.
\begin{itemize}
\item \textbf{Off-manifold guidance:} away from the data distribution, the vector field $-\nabla_x V_\theta$ is trained to align with transport-like directions that move samples from an easy-to-sample source distribution toward the data.
\item \textbf{Near-equilibrium refinement:} near equilibrium, $V_\theta$ acts as an EBM energy and defines a Gibbs density with unit temperature.
\end{itemize}
This separates learning into a \emph{global} transport-alignment term and a \emph{local} density refinement term.
A two-regime temperature $\epsilon(\cdot)$ is used: $\epsilon(t)=0$ off-manifold (transport regime) and $\epsilon(t)=1$ near equilibrium (mixing regime). 
This switch of regime (i)injects exploration, (ii) helps to prevent mode collapse, and (iii) stabilizes the Gibbs distribution in the mixing regime.

In continuous space, a JKO step for the energy-and-entropy functional reads
\begin{align}
\rho_{t+\Delta t}
&=
\arg\min_{\rho}\;
\frac{1}{2\Delta t}\,W_2^2(\rho,\rho_t)\nonumber\\
&\quad+\;\int V_\theta(x)\,\rho(x)\,\mathrm{d}x\nonumber\\
&\quad+\;\epsilon(t)\int \rho(x)\log\rho(x)\,\mathrm{d}x.
\end{align}

% \paragraph{Transport learning (off-manifold).}
\paragraph{Off-manifold guidance.}
These first-order optimality conditions in Wasserstein space~\citep{lanzetti2025firstorder,lanzetti2024variational} motivate two objectives~\citep{balcerak2025energy}:
(i) Away from data, transport directions are matched (ii) near equilibrium, the energy is refined to recover the Gibbs distribution.
Given optimally paired endpoints $(X_0,X_1)\sim\gamma$ with transport map $\gamma$ between the source distribution space $X_0\in \mathcal{X}_0$ and data distribution space $X_1\in \mathcal{X}_1$ and $t\sim\Unif([0,1])$, we define $X_t=(1-t)X_0+tX_1$ and
$v=X_1-X_0$. The potential $V_\theta$ is trained by matching the gradient field to $v$:
\begin{equation}
\mathcal{L}_{\mathrm{OT}}
=
\mathbb{E}_{(X_0,X_1)\sim\gamma,\;t\sim\Unif([0,1])}
\big[\|\nabla_x V_\theta(X_t) + v\|^2\big].
\end{equation}

% \paragraph{Energy refinement (near-equilibrium).}
\paragraph{Near-equilibrium refinement.}
As $t\to\infty$, the equilibrium distribution is
\begin{equation}
\rho_{t\to\infty}(x)\;\propto\;\exp\!\left(-V_\theta(x)\right).
\end{equation}
The refinement of $\rho_{t\to\infty}$ uses
\begin{equation}
\mathcal{L}_{\mathrm{ER}}
=
\mathbb{E}_{x^{+}\sim p_{\mathrm{data}}}\!\left[V_\theta(x^{+})\right]
-
\mathbb{E}_{x^{-}\sim \operatorname{sg}(p_{t\to\infty})}\!\left[V_\theta(x^{-})\right],
\end{equation}
where $x^{-}$ are sampled using Langevin dynamics, i.e., with a stop-gradient ($\operatorname{sg}$) through the sampler.
The sampling chain to generate these $x^{-}$ is initialized from the source or from data with a tunable mixing probability.
A warm-up phase with only
$\mathcal{L}_{\mathrm{OT}}$ allows source-initialized negative samples ($x^{-}$) to reach low-energy regions, followed byoptimization of
\begin{equation}
\mathcal{L}(\theta)=\mathcal{L}_{\mathrm{OT}}(\theta)+\lambda_{\mathrm{ER}}\,\mathcal{L}_{\mathrm{ER}}(\theta).
\end{equation}

\paragraph{Sampling (two regimes).}
Sampling starts from a transport-aligned drift and switches to a mixing via an effective temperature
$\epsilon(t)$. In continuous time, this is the time-inhomogeneous SDE
\begin{equation}
\mathrm{d}X_t
=
-\nabla_x V_\theta(X_t)\,\mathrm{d}t
+\sqrt{2\,\epsilon(t)}\,\mathrm{d}W_t,
\end{equation}
which reduces to an ODE when $\epsilon(t)=0$ (off-manifold) and becomes a stochastic dynamics as
the regime switches and $\epsilon(t)\to 1$. In practice, this dynamics needs to be integrated over time.

\paragraph{Related work: discrete diffusion/flow matching and control.}
Recent progress in discrete graph generation is dominated by denoising diffusion and CTMC
formulations. DiGress~\citep{vignac2022digress} extends discrete diffusion to categorical node and edge
attributes, while discrete CTMCs~\citep{campbell2024generative} move to a continuous-time graph diffusion process. Discrete
Flow Models (DFMs) show that flow-like probability paths on discrete spaces can be realized via
CTMCs~\citep{campbell2024generative}, and DeFoG~\citep{qinmadeira2024defog} instantiates this discrete
flow-matching viewpoint for graphs, emphasizing a more flexible separation between training and
sampling. Variational Flow Matching (VFM) recasts flow matching as variational inference over
trajectory endpoints, and CatFlow applies this to categorical data with strong results on graphs~\citep{eijkelboom2024vfm}.
For controllable generation, diffusion/CTMC guidance typically assumes access to a time- or
noise-dependent conditional signal along the path (e.g., a regressor/classifier for $p(y\mid x_t,t)$ or
$p_t(y\mid x_t)$). In discrete settings this often requires training a dedicated predictor on noised
latents across noise levels (or separate conditional/unconditional models for CFG), adding extra
training and coupling the control mechanism to the chosen noising process~\citep{schiff2024discreteguidance,eijkelboom2025controlledvfm}.
Moreover, property prediction on heavily corrupted graphs is ill-posed because bonds and valences are
often invalid, so regressor accuracy naturally degrades away from clean samples; we quantify this
effect in Appendix~\ref{app:regressor-noise}.
