\section{Experimental setup and hyperparameters}
\label{app:exp-setup}
\paragraph{MOSES GEM training (shared across experiments).}
Training is conducted in two stages. We first train with $\lambda_{\mathrm{CL}}=0.0$ for $330\,\text{k}$ iterations ($\text{lr}=10^{-4}$, batch size $=128$) to pretrain the potential field, enabling efficient burn-in of samples initialized from noise. We then follow with with
$\lambda_{\mathrm{CL}}=0.1$ for 250 iterations using chain length 500 at
$\text{lr}=5\times 10^{-5}$, batch size 128. L sampling uses greedy transport proposals followed by
dLangevin mixing with $\beta^L=9.55$, $\lambda_{\mathcal{V}}^L=0.23$, $\lambda_{\mathcal{E}}^L=1.88$,
and $\beta_{\mathrm{mh}}=\beta^L$. We incorporate RRWP features~\citep{ma2023graph} into GEM and all reproduced baselines to ensure a fair comparison with DeFoG, which originally employs these features. For the DeFoG baselines, we directly use the authors' provided checkpoint~\citep{qinmadeira2024defog}. During training, we consider two distinct setups for initializing the contrastive loss (CL) samples: either from noise or directly from data. We evenly split these initializations (50\% noise / 50\% data), except in the unconditional generation scenario (``Data Initialization'' category), where we exclusively (100\%) initialize CL samples from data.


\subsection{Unconditional generation}
We report 25k MOSES samples and 10k QM9 samples. For noise initialization we use greedy transport
followed by the mixing proposal with the parameters from the training setup. For data initialization we
use an annealed schedule with $\beta_{\mathrm{prop}}=8.12$, $\beta_{\mathrm{mh}}^{\mathrm{init}}=0.18$,
$\beta_{\mathrm{mh}}^{\mathrm{final}}=13.56$, $S_{\mathrm{anneal}}=200$, $\lambda_{\mathcal{V}}^L=0.07$,
and $\lambda_{\mathcal{E}}^L=2.23$.

\subsection{Conditional generation and property optimization}
\paragraph{Noisy property regressors.}
We train time-conditioned regressors on MOSES using shared hyper-parameters for both marginal- and uniform-noise variants. Training uses batch size $1024$ for $60$ epochs with Adam ($\text{lr}=10^{-4}$, weight decay=$10^{-6}$). The regressor is a GraphTransformer with $4$ layers, $8$ attention heads, hidden dimensions $(d_x,d_e,d_y)=(256,128,128)$, feedforward dimensions $(256,128,192)$, and MLP dimensions $(X,E,y)=(256,128,256)$, SiLU activations, and a scalar property head. We apply polydec time distortion and report test MAE on $10$k molecules at $t\in\{0,0.2,\dots,1\}$. Additional noise-sensitivity details are provided in Appendix~\ref{app:regressor-noise}.


\paragraph{Conditional generation (GEM \& DeFoG).}
GEM mixing proposal parameters:
($\beta^L=9.55$, $\lambda_{\mathcal{V}}^L=0.23$, $\lambda_{\mathcal{E}}^L=1.88$) with uniform noise initialization/regressors.
DeFoG uses marginal-noise initialization/ regressors. For each task we sweep $\lambda_{\text{prop}}$
and report the best CVUN; margins and selected $\lambda_{\text{prop}}$ are below. The reported results are computed directly using RDKit library tools—not from the regressors, which serve solely for guidance purposes.


\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lccc}
\toprule
Task & margin & GEM $\lambda_{\text{prop}}$ & DeFoG $\lambda_{\text{prop}}$ \\
\midrule
logP $\le 1.5$ & 0.2 & 10 & 140 \\
logS $\ge -2.25$ & 0.2 & 16 & 130 \\
QED $\ge 0.9$ & 0.05 & 193 & 360 \\
TPSA $\le 50$ & 5 & 0.03 & 0.5 \\
\bottomrule
\end{tabular}
\caption{\footnotesize Hyper-parameters for noisy-regressor conditioning on MOSES.
GEM uses uniform-noise regressors; DeFoG uses marginal-noise regressors. All results use 4k samples.}
\label{tab:cond-hparams}
\end{table}

\subsection{Geodesic analysis: validity along geodesics}
We compare GEM's energy-weighted geodesic to a cost-only baseline along interpolated paths between
matched MOSES molecules. We align permutations using the node-matching $c_{\mathrm{FGW}}$ cost
(Section~\ref{sec:gem}, Eq.~\ref{eq:fgw-cost}; node-matching details in Appendix~\ref{app:matching}),
and then use the aligned local cost $c_{\mathrm{loc}}$ (Eq.~\ref{eq:loc-cost}) to evaluate distances,
which is equivalent under a fixed alignment. Paths are parameterized as cubic B-splines in the
node/edge probability simplex with fixed endpoints and learned interior control points (8 control
points, degree 3). Optimization minimizes the energy-weighted path length $L_\theta$ from
Appendix~\ref{app:geodesic-details} (with the $\exp(\beta V_\theta)$ weighting) plus an average-energy
regularizer. We use $\beta=0.1$, length weight $\lambda_L=0.1$, energy weight $\lambda_{ME}=1.0$,
2000 Adam iterations, and a learning rate of $10^{-3}$ \citep{kingma2014adam}.

Validity is estimated by sampling 16 arc-length--uniform $s$ locations per path and drawing 16
discrete graphs per $s$. Validity is the fraction of RDKit-valid and connected samples, averaged
along each path and then across pairs within each distance bin. We use three distance bins:
$[5,10)$, $[10,15)$, and $[15,20)$, with 50 molecules per bin.
