\section{Permutation-invariant minibatch matching}
\label{app:matching}

Exact computation of the \gls{fgw} cost $c_{\mathrm{FGW}}$ requires a combinatorial search over
permutations. We therefore use fast, permutation-invariant hard-assignment approximations
tailored to the source distribution and the downstream use (pairing vs.\ local edits).

\paragraph{Histogram matching for a uniform source.}
When the source $\pi_0$ is a uniform distribution over node and edge classes, we use a cheap signature
for each graph. Let $h_\mathcal{V}$ be the normalized node-type histogram, $h_\mathcal{E}$ the normalized edge-type
histogram (over $i<j$), and $h_{\mathcal{V}\mathcal{E}}$ the histogram over unordered node-type pairs with edge types. We
form a weighted signature $h(x)=[\alpha_1 h_\mathcal{V},\;\alpha_2 h_\mathcal{E},\;\alpha_3 h_{\mathcal{V}\mathcal{E}}]$ and solve a linear assignment
between graphs of equal size using $\|h(x)-h(y)\|_1$. This produces permutation-invariant couplings and
is sufficient when $\pi_0$ is far from $\pi_{\mathrm{data}}$, so we do not require fine-grained alignment.

\paragraph{Node matching for molecule-like sources.}
If the source distribution already produces valid molecules, we recommend a node-matching alignment
based on node labels (optionally with a light edge-structure tie-breaker), followed by computing the
aligned local cost. This reflects chemical atom mapping practice: once atoms are aligned, bond
correspondences are largely determined~\citep{bell2019dockrmsd}.
