\section{Conceptualizing baselines}
\label{app:baselines}

We briefly summarize the baseline methods referenced in the experiments to clarify the modeling
assumptions behind each comparison.

\paragraph{GraphEBM.}
GraphEBM~\citep{liu2021graphebm} learns a scalar energy over molecular graphs with a GNN and relies on
MCMC (e.g., Metropolis-Hastings over local graph edits) for sampling, without an explicit transport
or flow-matching objective.

\paragraph{DeFoG.}
DeFoG~\citep{qinmadeira2024defog} formulates graph generation as discrete flow matching, learning
CTMC transition rates that connect source and data distributions; sampling follows the learned
continuous-time dynamics.

\paragraph{VFM.}
Variational Flow Matching (VFM)~\citep{eijkelboom2024vfm} reframes flow matching as variational
inference over trajectory endpoints, yielding globally consistent objectives for discrete graph
generation via CTMC paths.

\paragraph{G-VFM.}
G-VFM~\citep{eijkelboom2025controlledvfm} extends VFM with equivariant conditioning mechanisms to
support controllable generation, typically by incorporating property guidance into the flow-matching
framework.

\begin{table*}[t]
\centering
\small
\textbf{MOSES} ($\sim$1.5M molecules)\\[2pt]
\begingroup
\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{lcccccc}
\toprule
Method
& \multicolumn{2}{c}{\begin{tabular}{c}logP $>$ 1.5 \& logS $>$ 1.5\end{tabular}}
& \multicolumn{2}{c}{\begin{tabular}{c}logS $>$ 1.5 \& QED $>$ 0.6\end{tabular}}
& \multicolumn{2}{c}{\begin{tabular}{c}logP $>$ 1.5 \& QED $>$ 0.6\end{tabular}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
& C.V.U.N. $\uparrow$ & C.V.U. $\uparrow$
& C.V.U.N. $\uparrow$ & C.V.U. $\uparrow$
& C.V.U.N. $\uparrow$ & C.V.U. $\uparrow$ \\
\midrule
Training samples (MOSES) & 0 & 0 & 0 & 0 & 0 & 0 \\
\midrule
\multicolumn{7}{l}{\textit{Source Distribution Initialization}} \\
\midrule
GraphEBM~\citep{liu2021graphebm} & X & X & X & X & X & X \\
G-VFM~\citep{eijkelboom2025controlledvfm} & X & X & X & X & X & X \\
DeFoG$^*$~\citep{qinmadeira2024defog} & X & X & X & X & X & X \\
\textbf{GEM (Ours)} & X & X & X & X & X & X \\
\midrule
\multicolumn{7}{l}{\textit{Target Distribution Initialization}} \\
\midrule
GraphEBM~\citep{liu2021graphebm} & X & X & X & X & X & X \\
\textbf{GEM (Ours)} & \textbf{X} & \textbf{X} & \textbf{X} & \textbf{X} & \textbf{X} & \textbf{X} \\
\bottomrule
\end{tabular}
\endgroup
\vspace{0.15cm}
\caption{\footnotesize \textbf{Multi-Property Conditioning.} MOSES conditional generation with joint
constraints on (logP, logS) and (logS, QED). All entries are percentages for $C.V.U.N.$ and
$C.V.U.$. $^*$ reproduced. \textcolor{red}{[WiP]}}
\label{tab:moses-conditional-joint}
\end{table*}
