\section{Permutation-invariant energy parameterization}\label{app:perm-invariance}
We implement the potential with a graph transformer backbone directly from DeFoG~\cite{qinmadeira2024defog}. Each layer of this backbone is permutation equivariant, meaning that for any permutation (relabeling) $\sigma$, the layer satisfies:
\[
h^{(l)}(\sigma \cdot x_{\mathcal{V}}^{(l)}, \sigma \cdot x_{\mathcal{E}}^{(l)}, f^{(l)}) = (\sigma \cdot x_{\mathcal{V}}^{(l+1)}, \sigma \cdot x_{\mathcal{E}}^{(l+1)}, f^{(l+1)}),
\]
where $x_{\mathcal{V}}^{(l)}$ and $x_{\mathcal{E}}^{(l)}$ denote node and edge features at layer $l$, respectively, and $f^{(l)}$ denotes the global graph-level feature.

Following~\citep{vignac2022digress,qinmadeira2024defog}, we pad and mask all smaller graphs to the maximum number of nodes observed in the training dataset, denoted as $n$. This results in fixed-size inputs to the model, denoted $x_\mathcal{V}$ and $x_\mathcal{E}$ for node and edge features, respectively. We denote node and edge features at layer $l$ as $x^{(l)}_\mathcal{V}$ and $x^{(l)}_\mathcal{E}$, respectively.
The global feature $f^{(l)}$ collects graph-level features at layer $l$ and is updated through a permutation-invariant operation.
Figure~\ref{fig:energy-arch} summarizes the backbone and invariant readout.


\begin{figure}[h]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
  font=\scriptsize,
  node distance=1.0cm,
  module/.style={
    rectangle,
    rounded corners=5pt,
    minimum width=3.1cm,
    minimum height=0.95cm,
    align=center,
    draw=black!55,
    fill=black!4,
    thick
  },
  arrow/.style={-Latex, line width=1.1pt, draw=black!60}
]
  \node[module] (graph) {Input Graph \\\\ $x:=(x^{(l)}_\mathcal{V},x^{(l)}_\mathcal{E})$};
  \node[module, right=of graph] (gt) {Graph Transformer\\(Permutation Equivariant)};
  \node[module, right=of gt] (pool) {Invariant Readout\\\\$\sum_i x^{(L)}_i,\;\sum_{i\ne j} x^{(L)}_{ij},f^{(L)}$ };
  \node[module, right=of pool] (energy) {$V_\theta(x)\in\mathbb{R}$};

  \draw[arrow] (graph) -- (gt);
  \draw[arrow] (gt) -- (pool);
  \draw[arrow] (pool) -- (energy);
\end{tikzpicture}}
\caption{\textbf{Energy Backbone.} Permutation-equivariant backbone with invariant pooling to produce a scalar energy.}
\label{fig:energy-arch}
\end{figure}

The scalar energy is obtained by an invariant readout that first pools the
equivariant features and then applies an \gls{mlp}:
\begin{equation}
\begin{aligned}
\bar{x} &= \sum_{i\in\mathcal{V}} x^{(L)}_i,\qquad
\bar{e} = \sum_{i< j} x^{(L)}_{ij},\\
V_\theta(x) &= \mathrm{MLP}\big([\bar{x},\bar{e},f^{(L)}]\big).
\end{aligned}
\end{equation}
Since the sum operation is permutation invariant, the concatenated vector
$(\bar{x},\bar{e},f^{(L)})$ is invariant. Therefore, an \gls{mlp} applied to it is
also invariant.

\paragraph{Energy head and parameter count.}
To enhance expressivity, we incorporate Relative Random Walk Probabilities (RRWP)~\citep{ma2023graph}
as node and edge features, matching DeFoG for a fair comparison. The energy MLP head is a 2-layer MLP
that maps the pooled hidden graph representation (concatenated node/edge/$y$ embeddings) to a single
scalar energy. With dimensions $448\to256\to1$ and SiLU activation, it has 115,201 parameters. The
full model has 16,382,158 trainable parameters ($\approx 16.38$M).
